Connecting the Popularity Adjusted Block Model to the Generalized Random Dot Product Graph for Community Detection and Parameter Estimation

### 2021-03-22 summary.Rmd
fixed theorem 4

---
title: "Connecting the Popularity Adjusted Block Model to the Generalized Random Dot Product Graph for Community Detection and Parameter Estimation"
author: John Koo, Minh Tang, Michael Trosset
output:
  pdf_document:
    citation_package: natbib
    number_sections: yes
# output: html_document
fontsize: 12pt
geometry: "left=1in,right=1in,top=1in,bottom=1in"
urlcolor: blue
header-includes:
- \usepackage{float}
- \usepackage{mathtools}
- \usepackage{natbib}
- \usepackage[linesnumbered,ruled,vlined]{algorithm2e} 
- \setcitestyle{numbers,square}
- \usepackage{verbatim}
bibliography: misc.bib
abstract: |
  In this paper, we connect two random graph models, the Popularity 
  Adjusted Block Model (PABM) and the Generalized Random Dot Product Graph 
  (GRDPG) and use properties established in this connection to aid in 
  community detection and parameter estimation. In particular, we note that the
  PABM can be represented as latent positions such that points within the same 
  community lie on a subspace, and the subspaces that represent each 
  community are orthogonal to one another. Using this property as well as the 
  asymptotic properties of Adjacency Spectral Embedding (ASE) of the GRDPG, we 
  are able to establish theoretical asymptotic results of our community 
  detection and parameter estimation methods for the PABM, including for 
  Sparse Subspace Clustering (SSC). 
---

@@

\newcommand{\diag}{\text{diag}}
\newcommand{\tr}{\text{Tr}}
\newcommand{\blockdiag}{\text{blockdiag}}
\newcommand{\indep}{\stackrel{\text{indep}}{\sim}}
\newcommand{\iid}{\stackrel{\text{iid}}{\sim}}
\newcommand{\Bernoulli}{\text{Bernoulli}}
\newcommand{\Betadist}{\text{Beta}}

@@

**Definition 1** (Popularity Adjusted Block Model) 
\cite{307cbeb9b1be48299388437423d94bf1}. 
Let $P \in [0, 1]^{n \times n}$ be a symmetric edge probability matrix for a 
set of $n$ 
vertices, $V$. Each vertex has a community label $1, ..., K$, and the rows and 
columns of $P$ are arranged by community label such that $n_k \times n_l$ block 
$P^{(kl)}$ describes the edge probabilities between vertices in communities 
$k$ and $l$ ($P^{(lk)} = (P^{(kl)})^\top$). 
Let graph $G = (V, E)$ be an undirected, unweighted graph such 
that its corresponding adjacency matrix $A \in \{0, 1\}^{n \times n}$ is a 
realization of $\Bernoulli(P)$, i.e., 
$A_{ij} \indep \Bernoulli(P_{ij})$ for $i > j$ 
($A_{ij} = A_{ji}$ and $A_{ii} = 0$).

@@

for a set of $K^2$ fixed vectors $\{\lambda^{(st)}\}_{s, t = 1}^K$ where each 
$\lambda^{(st)}$ is a column vector 
of dimension $n_s$, then graph $G$ and its corresponding adjacency matrix $A$ 
is a realization of a popularity adjusted block model with parameters 
$\{\lambda^{(st)}\}_{s, t = 1}^K$.

@@

**Definition 2** (Generalized Random Dot Product Graph) 
\cite{rubindelanchy2017statistical}.
Let $P \in [0, 1]^{n \times n}$ be a symmetric edge probability matrix for a 
set of $n$ vertices, $V$. If $\exists X \in \mathbb{R}^{n \times d}$ such that

@@

for some $d, p, q \in \mathbb{N}$ and $p + q = d$, then 
graph $G = (V, E)$ with adjacency matrix $A$ such that 
$A_{ij} \indep \Bernoulli(P_{ij})$ for $i > j$ ($A_{ij} = A_{ji}$ and 
$A_{ii} = 0$) is a draw from the generalized random dot product graph model 
with latent positions $X$ and signature $(p, q)$. More precisely, if vertices 
$i$ and $j$ have latent positions $x_i$ and $x_j$ respectively, then the edge 
probability between the two is $P_{ij} = x_i^\top I_{pq} x_j$, and $X$ contains 
the latent positions as rows $x_i^\top$.

@@

**Definition 3**. The indefinite orthogonal group with signature $(p, q)$ is 
the set $\{Q \in \mathbb{R}^{d \times d} : Q I_{pq} Q^{\top} = I_{pq}\}$, 
denoted as $\mathbb{O}(p, q)$ \cite{rubindelanchy2017statistical}.

**Remark**. 
Like the RDPG, the latent positions of a GRDPG are not unique 
\cite{rubindelanchy2017statistical}. 
More specifically, if $P_{ij} = x_i^\top I_{pq} x_j$, then we also have for any 
$Q \in \mathbb{O}(p, q)$, 
$(Q x_i)^\top I_{pq} (Q x_j) = x_i^\top (Q^\top I_{pq} Q) x_j = 
x_i^\top I_{pq} x_j = P_{ij}$. 
Unlike in the RDPG case, transforming the latent positions via multiplication 
by $Q \in \mathbb{O}(p, q)$ does not necessarily maintain interpoint angles or 
distances.

@@

**Theorem 1** (Connecting the PABM to the GRDPG for $K = 2$). 
Let

@@

where each $\lambda^{(kl)}$ is a vector as in Definition 1. 
Then $A \sim GRDPG_{3, 1}(X U)$ and $A \sim PABM(\{(\lambda^{(kl)}\}_2)$ are 
equivalent.

**Theorem 2** (Generalization to $K > 2$). There exists a block diagonal matrix 
$X \in \mathbb{R}^{n \times K^2}$ defined by PABM parameters 
$\{\lambda^{(kl)}\}_K$ and orthonormal matrix 
$U \in \mathbb{R}^{K^2 \times K^2}$ that is fixed 
for each $K$ such that $A \sim GRDPG_{K (K+1) / 2, K (K-1) / 2}(XU)$ and 
$A \sim PABM(\{(\lambda^{(kl)}\})_K)$ are equivalent.

*Proof of Theorem 2*. 
Define the following matrices from $\{\lambda^{(kl)}\}_K$: 
$$\Lambda^{(k)} = 
\begin{bmatrix} \lambda^{(k,1)} & \cdots & \lambda^{(k, K)} \end{bmatrix}
\in \mathbb{R}^{n_k \times K}$$

@@

and we can describe the PABM with $K$ communities as a GRDPG with latent 
positions $X U$ with signature $\big( K (K + 1) / 2, K (K - 1) / 2 \big)$.

**Example** ($K = 3$). Using the same notation as before:

@@

* Positions 1, 5, 9 are fixed.
* The cycles of order 2 are $(2, 4)$, $(3, 7)$, and $(6, 8)$.

@@

The latent positions are the rows of  
$$XU = \begin{bmatrix}
  \lambda^{(11)} & 0 & 0 & 
  \lambda^{(12)} / \sqrt{2} & \lambda^{(13)} / \sqrt{2} & 0 & 
  \lambda^{(12)} / \sqrt{2} & \lambda^{(13)} / \sqrt{2} & 0 \\
  0 & \lambda^{(22)} & 0 & 
  \lambda^{(21)} / \sqrt{2} & 0 & \lambda^{(23)} / \sqrt{2} & 
  -\lambda^{(21)} / \sqrt{2} & 0 & \lambda^{(23)} / \sqrt{2} \\
  0 & 0 & \lambda^{(33)} & 
  0 & \lambda^{(31)} / \sqrt{2} & \lambda^{(32)} / \sqrt{2} & 
  0 & -\lambda^{(31)} / \sqrt{2} & -\lambda^{(32)} / \sqrt{2}
\end{bmatrix}$$

@@

**Definition 4** (Subspace Detection Property) \cite{pmlr-v28-wang13}. 
Let $X = \begin{bmatrix} x_1 & \cdots & x_n \end{bmatrix}^\top$ be noisy 
points sampled from $K$ subspaces. Let $C$ and $B$ be constructed from the 
solutions of LASSO problems as described in (\ref{eq:ssc}). If each column of 
$C$ has nonzero norm and $B_{ij} = 0$ $\forall$ $x_i$ and $x_j$ sampled from 
different subspaces, then $X$ obeys the subspace detection property.

**Remark**. In practice, sample $X$ often does not obey the subspace detection
property. In such cases, $B$ is treated as an affinity matrix for a graph which 
is then partitioned into $K$ subgraphs to obtain the clustering. On the other 
hand, if $X$ does obey the subspace detection property, $B$ describes a graph 
with at least $K$ disconnected subgraphs. Ideally, when the subspace detection
property holds, there are exactly $K$ subgraphs which map to each subspace, 
but it could be the case that some of the subspaces are represented by 
multiple disconnected subgraphs.

Theorem 2 suggests that SSC is appropriate for community detection for the 
PABM. More precisely, Theorem 2 says that each community consists of a
$K$-dimensional subspace, and together the subspaces lie in $\mathbb{R}^{K^2}$. 
The natural approach then is to perform SSC on the ASE of $P$ or $A$.
\citet{noroozi2019estimation} instead applied SSC to $P$ and $A$, foregoing 
embedding altogether.

Using results from \citet{soltanolkotabi2012}, it can be easily shown that 
the subspace detection property holds for $XU$, which is an ASE of $P$. More 
specifically, if points lie exactly on mutually orthogonal subspaces, then the 
subspace detection property will hold with probability 1, and this is exactly 
the case for the PABM (Theorem 2). Much of our work is then built on \citet{rubindelanchy2017statistical}, who describe the convergence behavior of 
the ASE of $A$ to the ASE of $P$, and \citet{pmlr-v28-wang13}, who show the 
necessary conditions for the subspace detection property to hold in noisy cases 
where the points lie near subspaces.

@@

\begin{algorithm}[t]
  \DontPrintSemicolon
  \SetAlgoLined
  \KwData{Adjacency matrix $A$, number of communities $K$}
  \KwResult{Community assignments $1, ..., K$}
    Compute the eigenvectors of $A$ that correspond to the $K (K+1) / 2$ most 
    positive eigenvalues and $K (K-1) / 2$ most negative eigenvalues. Construct 
    $V$ using these eigenvectors as its columns.\;
    Compute $B = |V V^\top|$, applying $|\cdot|$ entry-wise.\;
    Construct graph $G$ using $B$ as its similarity matrix.\;
    Partition $G$ into $K$ disconnected subgraphs  
    (e.g., using edge thresholding or spectral clustering).\;
    Map each partition to the community labels $1, ..., K$.\;
  \caption{Orthogonal Spectral Clustering.}
\end{algorithm}

@@

**Theorem 3.** 
Let $P = V D V^\top$ be the spectral decomposition of the edge probability
matrix. Let $B = V V^\top$. Then $B_{ij} = 0$ if vertices $i$ and $j$ are 
from different communities.

Theorem 3 provides perfect community detection given $P$. Letting $|B|$ be the 
affinity matrix for graph $G$, $G$ is partitioned into $K$ disjoint subgraphs 
that correspond to each community. Using $A$ introduces some error in the ASE, 
resulting in points that lie near subspaces but not exactly on them. Using 
results from \cite{rubindelanchy2017statistical}, we can show that as $n$ 
grows, the ASE of $A$ approaches the ASE of $P$ and $B_{ij} \to 0$ for $i$ and 
$j$ in different clusters.

The ASE of $A$ approaches latent positions that form $P$ as 
the number of vertices $n$ increases. More precisely, let each 
$\lambda^{(kl)}_i \sim \mathcal{F}^{(kl)}$ for $i = 1, .., n$ and 
$k, l = 1, ..., K$. Then the corresponding latent positions 
$XU \sim \mathcal{G}_K$ for some related joint distribution with $K$ underlying
communities $\mathcal{G}_K$. Denote $Z_n$ as a sample of size $n$ from
$\mathcal{G}_K$ and adjacency matrix $A_n$ as one draw from edge probability 
matrix $P_n = Z_n I_{pq} Z_n^\top$. Let $\hat{Z}_n$ be the adjacency embedding 
of $A_n$ with rows $(\hat{z}_i^{(n)})^\top$. Then by
\citet{rubindelanchy2017statistical},

\begin{equation} \label{eq:thm3}
  \max\limits_{i \in \{1, ..., n\}} ||Q_n \hat{z}_i^{(n)} - z_i^{(n)}|| = 
  O_P\bigg(\frac{(\log n)^c}{n^{1/2}} \bigg)
\end{equation}

for some $c > 0$ and sequence of $Q_n \in \mathbb{O}(p, q)$.

**Theorem 4**. 
Let $\hat{B}_n$ with entries $\hat{B}_n^{(ij})$ be the affinity matrix from OSC (Alg. 1). Then $\forall$ pairs $(i, j)$ belonging to different communities,

\begin{equation} \label{eq:thm4}
\hat{B}_n^{(ij)} \stackrel{a.s.}{\to} 0
\end{equation}

Theorems 2, 3, and 4 also provide a very natural path toward using SSC for 
community detection for the PABM. We established in Theorem 2 that an ASE of 
the edge probability matrix $P$ can be constructed such that the communities 
lie on mutually orthogonal subspaces, and this property can be recovered from
the eigenvectors of $P$. Then Theorems 3 and 4 show that this property holds 
for the adjacency matrix $A$ drawn from $P$ as $n \to \infty$.

\begin{algorithm}[t]
  \DontPrintSemicolon
  \SetAlgoLined
  \caption{Sparse Subspace Clustering using LASSO \cite{pmlr-v28-wang13}.}
  \KwData{Adjacency matrix $A$, number of communities $K$, 
  hyperparameter $\lambda$}
  \KwResult{Community assignments $1, ..., K$}
    Find $Z$, the ASE of $A$ using the $K (K + 1) / 2$ most positive 
    and the $K (K - 1) / 2$ most negative eigenvalues and eigenvectors of 
    $A$.\;
    \For {$i = 1, ..., n$} {
      Assign $z_i^\top$ as the $i^{th}$ row of $Z$. 
      Assign $Z_{-i} = \begin{bmatrix} 
      z_1 & \cdots & z_{i-1} & z_{i+1} & \cdots & z_n \end{bmatrix}^\top$.\;
      Solve the LASSO problem 
      $c_i = \arg\min_{\beta_i} 
      \frac{1}{2} ||z_i - Z_{-i} \beta_i||_2^2 + \lambda ||\beta_i||_1$.\;
      Assign $\tilde{c}_i = \begin{bmatrix} 
      c_i^{(1)} & \cdots & c_i^{(i-1)} & 0 & c_i^{(i)} & \cdots & c_i^{(n-1)}
      \end{bmatrix}^\top$ such that the superscript is the index of 
      $\tilde{c}_i$.\;
    }
    Assign 
    $C = \begin{bmatrix} \tilde{c}_1 & \cdots & \tilde{c}_n \end{bmatrix}$.\;
    Compute the affinity matrix $B = |C| + |C^\top|$.\;
    Construct graph $G$ using $B$ as its similarity matrix.\;
    Partition $G$ into $K$ disconnected subgraphs (e.g., using edge 
    thresholding or spectral clustering).\;
    Map each partition to the community labels $1, ..., K$. 
\end{algorithm}

**Theorem 5**. Let $P_n$ describe the edge probability matrix of the PABM with 
$n$ vertices, and let $A_n \sim \Bernoulli(P_n)$.  Let $\hat{Z}_n$ be the ASE 
of $A_n$. Then $\exists \lambda > 0$ such that $\hat{B}_n$, the affinity matrix
outputted by SSC using $\hat{Z}_n$ and $\lambda$, has $ij$^th^ entry
$\hat{B}_n^{(ij)} \stackrel{a.s.}{\to} 0$ $\forall i, j$ in different 
communities.

@@

For any edge probability matrix $P$ for the PABM such that the rows and 
columns are organized by community, the $kl$^th^ block is an outer product 
of two vectors, i.e., $P^{(kl)} = \lambda^{(kl)} (\lambda^{(lk)})^\top$. 
Therefore, given $P^{(kl)}$, $\lambda^{(kl)}$ and $\lambda^{(lk)}$ are 
solvable exactly (up to multiplication by $-1$) using singular value 
decomposition. More specifically, let $P^{(kl)} = \sigma^2 u v^\top$ be the 
singular value decomposition of $P^{(kl)}$. $u \in \mathbb{R}^{n_k}$ and 
$v \in \mathbb{R}^{n_l}$ are vectors and 
$\sigma^2 > 0$ is a scalar. Then $\lambda^{(kl)} = \pm \sigma u$ and 
$\lambda^{(lk)} = \pm \sigma v$. 
Given the adjacency matrix $A$ instead of edge probability matrix $P$, we can 
simply use plug-in estimators (algorithm 3), which converge to the true 
parameters (Theorem 6).

@@

**Theorem 6**. Under regularity and sparsity assumptions, and under the further 
assumption that $K$ is fixed and community labels are known,

\begin{equation} \label{eq:thm6}
\max_{k, l \in \{1, ..., K\}} 
||\hat{\lambda}^{(kl)} - \lambda^{(kl)}|| = 
O_P \bigg(\frac{(\log n_k)^c}{\sqrt{n_k}} \bigg)
\end{equation}

@@

For each simulation, community labels are drawn from a multinomial 
distribution, the popularity vectors $\{\lambda^{(kl)}\}_K$ are drawn 
from two types of joint distributions depending on whether $k = l$, the edge
probability matrix $P$ is constructed using 
the popularity vectors, and finally an unweighted and undirected adjacency 
matrix $A$ is drawn from $P$. OSC is then used for community detection, 
and this method is compared against SSC \cite{noroozi2019estimation}
\cite{soltanolkotabi2014} and MM \cite{igraph}
\cite{307cbeb9b1be48299388437423d94bf1}. True 
community labels are used with Algorithm 3 to estimate the popularity 
vectors $\{\lambda^{(kl)}\}_K$, and this method is then compared against an
MLE-based estimator described in \citet{noroozi2019estimation} and
\citet{307cbeb9b1be48299388437423d94bf1}.

@@

Theorem 4 implies that OSC will result in not just in the error rate 
converging to 0 but the error *count* as well.
We explore this in Fig. \ref{fig:clust_err_ct_sim}.

>>>>

---
title: "Connecting the Popularity Adjusted Block Model to the Generalized Random Dot Product Graph for Community Detection and Parameter Estimation"
authors: John Koo, Minh Tang, Michael Trosset
# author: John Koo
# author2: Minh Tang
# author3: Michael Trosset
output:
  pdf_document:
    citation_package: natbib
    number_sections: yes
# output:
#   bookdown::pdf_document2:
#     citation_package: natbib
#     number_citations: yes
# output: html_document
# output: rticles::rss_article
fontsize: 12pt
# geometry: "left=1in,right=1in,top=1in,bottom=1in"
urlcolor: blue
header-includes:
- \usepackage{float}
- \usepackage{mathtools}
- \usepackage{natbib}
- \usepackage[linesnumbered,ruled,vlined]{algorithm2e} 
- \setcitestyle{numbers,square}
- \usepackage{verbatim}
- \usepackage{amsthm}
bibliography: misc.bib
abstract: |
  We connect two random graph models, the Popularity Adjusted Block Model 
  (PABM) and the Generalized Random Dot Product Graph (GRDPG), demonstrating
  that a PABM is a GRDPG in which communities correspond to certain mutually
  orthogonal subspaces of latent positions. This insight leads to the
  construction of improved algorithms for community detection and parameter
  estimation with PABM. Using established asymptotic properties of Adjacency
  Spectral Embedding (ASE) for GRDPG, we derive asymptotic properties of these
  algorithms, including algorithms that rely on Sparse Subspace Clustering
  (SSC). We illustrate these properties via simulation.
---

@@

\providecommand{\tightlist}{%
  \setlength{\itemsep}{0pt}\setlength{\parskip}{0pt}}

\newcommand{\diag}{\text{diag}}
\newcommand{\tr}{\text{Tr}}
\newcommand{\blockdiag}{\text{blockdiag}}
\newcommand{\indep}{\stackrel{\text{indep}}{\sim}}
\newcommand{\iid}{\stackrel{\text{iid}}{\sim}}
\newcommand{\Bernoulli}{\text{Bernoulli}}
\newcommand{\Betadist}{\text{Beta}}
\newtheorem{definition}{Definition}
\newtheorem{theorem}{Theorem}
\newtheorem{lemma}{Lemma}
\theoremstyle{remark}
\newtheorem*{remark}{Remark}
\theoremstyle{example}
\newtheorem*{example}{Example}

@@

\begin{definition}[Popularity Adjusted Block Model]
\label{pabm}
\cite{307cbeb9b1be48299388437423d94bf1}. 
Let $P \in [0, 1]^{n \times n}$ be a symmetric edge probability matrix for a 
set of $n$ 
vertices, $V$. Each vertex has a community label $1, ..., K$, and the rows and 
columns of $P$ are arranged by community label such that $n_k \times n_l$ block 
$P^{(kl)}$ describes the edge probabilities between vertices in communities 
$k$ and $l$ ($P^{(lk)} = (P^{(kl)})^\top$). 
Let graph $G = (V, E)$ be an undirected, unweighted graph such 
that its corresponding adjacency matrix $A \in \{0, 1\}^{n \times n}$ is a 
realization of $\Bernoulli(P)$, i.e., 
$A_{ij} \indep \Bernoulli(P_{ij})$ for $i > j$ 
($A_{ij} = A_{ji}$ and $A_{ii} = 0$).

@@

for a set of $K^2$ fixed vectors $\{\lambda^{(st)}\}_{s, t = 1}^K$ where each 
$\lambda^{(st)}$ is a column vector 
of dimension $n_s$, then graph $G$ and its corresponding adjacency matrix $A$ 
is a realization of a popularity adjusted block model with parameters 
$\{\lambda^{(st)}\}_{s, t = 1}^K$. 
\end{definition}

@@

\begin{definition}[Generalized Random Dot Product Graph] 
\label{grdpg} 
\cite{rubindelanchy2017statistical}.
Let $P \in [0, 1]^{n \times n}$ be a symmetric edge probability matrix for a 
set of $n$ vertices, $V$. If $\exists X \in \mathbb{R}^{n \times d}$ such that

@@

for some $d, p, q \in \mathbb{N}$ and $p + q = d$, then 
graph $G = (V, E)$ with adjacency matrix $A$ such that 
$A_{ij} \indep \Bernoulli(P_{ij})$ for $i > j$ ($A_{ij} = A_{ji}$ and 
$A_{ii} = 0$) is a draw from the generalized random dot product graph model 
with latent positions $X$ and signature $(p, q)$. More precisely, if vertices 
$i$ and $j$ have latent positions $x_i$ and $x_j$ respectively, then the edge 
probability between the two is $P_{ij} = x_i^\top I_{pq} x_j$, and $X$ contains 
the latent positions as rows $x_i^\top$. 
\end{definition}

@@

\begin{definition}[Indefinite Orthogonal Group] 
The indefinite orthogonal group with signature $(p, q)$ is 
the set $\{Q \in \mathbb{R}^{d \times d} : Q I_{pq} Q^{\top} = I_{pq}\}$, 
denoted as $\mathbb{O}(p, q)$ \cite{rubindelanchy2017statistical}.
\end{definition}

\begin{remark}
Like the RDPG, the latent positions of a GRDPG are not unique 
\cite{rubindelanchy2017statistical}. 
More specifically, if $P_{ij} = x_i^\top I_{pq} x_j$, then we also have for any 
$Q \in \mathbb{O}(p, q)$, 
$(Q x_i)^\top I_{pq} (Q x_j) = x_i^\top (Q^\top I_{pq} Q) x_j = 
x_i^\top I_{pq} x_j = P_{ij}$. 
Unlike in the RDPG case, transforming the latent positions via multiplication 
by $Q \in \mathbb{O}(p, q)$ does not necessarily maintain interpoint angles or 
distances. 
\end{remark}

@@

\begin{theorem}[Connecting the PABM to the GRDPG for $K = 2$]
\label{theorem1}  
Let

@@

where each $\lambda^{(kl)}$ is a vector as in Definition 1. 
Then $A \sim GRDPG_{3, 1}(X U)$ and $A \sim PABM(\{(\lambda^{(kl)}\}_2)$ are 
equivalent.
\end{theorem}

\begin{theorem}[Generalization to $K > 2$] 
\label{theorem2}
There exists a block diagonal matrix 
$X \in \mathbb{R}^{n \times K^2}$ defined by PABM parameters 
$\{\lambda^{(kl)}\}_K$ and orthonormal matrix 
$U \in \mathbb{R}^{K^2 \times K^2}$ that is fixed 
for each $K$ such that $A \sim GRDPG_{K (K+1) / 2, K (K-1) / 2}(XU)$ and 
$A \sim PABM(\{(\lambda^{(kl)}\})_K)$ are equivalent.
\end{theorem}

\begin{proof}
Define the following matrices from $\{\lambda^{(kl)}\}_K$: 
$$\Lambda^{(k)} = 
\begin{bmatrix} \lambda^{(k,1)} & \cdots & \lambda^{(k, K)} \end{bmatrix}
\in \mathbb{R}^{n_k \times K}$$

@@

and we can describe the PABM with $K$ communities as a GRDPG with latent 
positions $X U$ with signature $\big( K (K + 1) / 2, K (K - 1) / 2 \big)$.
\end{proof}

\begin{example}[$K = 3$] Using the same notation as in Theorem \ref{theorem2}:

@@

* Positions 1, 5, 9 are fixed.

* The cycles of order 2 are $(2, 4)$, $(3, 7)$, and $(6, 8)$.

@@

The latent positions are the rows of  
$$XU = \begin{bmatrix}
  \lambda^{(11)} & 0 & 0 & 
  \lambda^{(12)} / \sqrt{2} & \lambda^{(13)} / \sqrt{2} & 0 & 
  \lambda^{(12)} / \sqrt{2} & \lambda^{(13)} / \sqrt{2} & 0 \\
  0 & \lambda^{(22)} & 0 & 
  \lambda^{(21)} / \sqrt{2} & 0 & \lambda^{(23)} / \sqrt{2} & 
  -\lambda^{(21)} / \sqrt{2} & 0 & \lambda^{(23)} / \sqrt{2} \\
  0 & 0 & \lambda^{(33)} & 
  0 & \lambda^{(31)} / \sqrt{2} & \lambda^{(32)} / \sqrt{2} & 
  0 & -\lambda^{(31)} / \sqrt{2} & -\lambda^{(32)} / \sqrt{2}
\end{bmatrix}$$
\end{example}

@@

\begin{definition}[Subspace Detection Property \cite{pmlr-v28-wang13}]. 
Let $X = \begin{bmatrix} x_1 & \cdots & x_n \end{bmatrix}^\top$ be noisy 
points sampled from $K$ subspaces. Let $C$ and $B$ be constructed from the 
solutions of LASSO problems as described in (\ref{eq:ssc}). If each column of 
$C$ has nonzero norm and $B_{ij} = 0$ $\forall$ $x_i$ and $x_j$ sampled from 
different subspaces, then $X$ obeys the subspace detection property. 
\end{definition}

\begin{remark} 
In practice, sample $X$ often does not obey the subspace detection
property. In such cases, $B$ is treated as an affinity matrix for a graph which 
is then partitioned into $K$ subgraphs to obtain the clustering. On the other 
hand, if $X$ does obey the subspace detection property, $B$ describes a graph 
with at least $K$ disconnected subgraphs. Ideally, when the subspace detection
property holds, there are exactly $K$ subgraphs which map to each subspace, 
but it could be the case that some of the subspaces are represented by 
multiple disconnected subgraphs. 
\end{remark}

Theorem \ref{theorem2} suggests that SSC is appropriate for community 
detection for the PABM. 
More precisely, Theorem \ref{theorem2} says that each community consists of a
$K$-dimensional subspace, and together the subspaces lie in $\mathbb{R}^{K^2}$. 
The natural approach then is to perform SSC on the ASE of $P$ or $A$.
\citeauthor{noroozi2019estimation} instead applied SSC to $P$ and $A$, 
foregoing embedding altogether.

Using results from \citeauthor{soltanolkotabi2012}, it can be easily shown that 
the subspace detection property holds for $XU$, which is an ASE of $P$. More 
specifically, if points lie exactly on mutually orthogonal subspaces, then the 
subspace detection property will hold with probability 1, and this is exactly 
the case for the PABM (Theorem \ref{theorem2}). 
Much of our work is then built on
\citeauthor{rubindelanchy2017statistical}, who describe the convergence 
behavior of the ASE of $A$ to the ASE of $P$, and \citeauthor{pmlr-v28-wang13},
who show the necessary conditions for the subspace detection property to hold 
in noisy cases where the points lie near subspaces.

@@

\begin{algorithm}[t]
  \DontPrintSemicolon
  \SetAlgoLined
  \KwData{Adjacency matrix $A$, number of communities $K$}
  \KwResult{Community assignments $1, ..., K$}
    Compute the eigenvectors of $A$ that correspond to the $K (K+1) / 2$ most 
    positive eigenvalues and $K (K-1) / 2$ most negative eigenvalues. Construct 
    $V$ using these eigenvectors as its columns.\;
    Compute $B = |n V V^\top|$, applying $|\cdot|$ entry-wise.\;
    Construct graph $G$ using $B$ as its similarity matrix.\;
    Partition $G$ into $K$ disconnected subgraphs  
    (e.g., using edge thresholding or spectral clustering).\;
    Map each partition to the community labels $1, ..., K$.\;
  \caption{Orthogonal Spectral Clustering.}
\end{algorithm}

@@

\begin{theorem}
\label{theorem3}
Let $P = V D V^\top$ be the spectral decomposition of the edge probability
matrix. Let $B = n V V^\top$. Then $B_{ij} = 0$ if vertices $i$ and $j$ are 
from different communities.
\end{theorem}

Theorem \ref{theorem3} provides perfect community detection given $P$. 
Letting $|B|$ be the affinity matrix for graph $G$, $G$ is partitioned into 
at least $K$ disjoint subgraphs since each of the $K$ communities have no 
edges between them. Similar to the subspace detection property, it could be 
the case that some of the communities are represented by multiple disjoint
subgraphs in $G$, in which case additional reconstruction is required to
identify the communities exactly.

Using $A$ instead of $P$ introduces error, which converges to $0$ 
almost surely:

\begin{theorem}
\label{theorem4} 
Let $\hat{B}_n$ with entries $\hat{B}_n^{(ij)}$ be the affinity matrix from OSC 
(Alg. 1). Then $\forall$ pairs $(i, j)$ belonging to different communities 
and sparsity factor satisfying $n \rho_n = \omega\{(\log n)^{4c}\}$,

\begin{equation} \label{eq:thm4}
\max_{i, j} |n (\hat{v}_n^{(i)})^\top \hat{v}_n^{(j)}| = 
O_P \Big( \frac{(\log n)^c}{\sqrt{n \rho_n}} \Big)
\end{equation}

This provides the result that for $i, j$ in different communities, 
$\hat{B}_n^{(ij)} \stackrel{a.s.}{\to} 0$.
\end{theorem}

Theorems \ref{theorem2}, \ref{theorem3}, and \ref{theorem4} 
also provide a very natural path toward using SSC for 
community detection for the PABM. We established in Theorem \ref{theorem2}
that an ASE of the edge probability matrix $P$ can be constructed 
such that the communities lie on mutually orthogonal subspaces,
and this property can be recovered from the eigenvectors of $P$. 
Then Theorems \ref{theorem3} and \ref{theorem4} show that this property holds 
for the unscaled ASE of $A$ drawn from $P$ as $n \to \infty$.

\begin{algorithm}[t]
  \DontPrintSemicolon
  \SetAlgoLined
  \caption{Sparse Subspace Clustering using LASSO \cite{pmlr-v28-wang13}.}
  \KwData{Adjacency matrix $A$, number of communities $K$, 
  hyperparameter $\lambda$}
  \KwResult{Community assignments $1, ..., K$}
    Find $V$, the matrix of eigenvectors of $A$ 
    corresponding to the $K (K + 1) / 2$ most positive 
    and the $K (K - 1) / 2$ most negative eigenvalues.\;
    \For {$i = 1, ..., n$} {
      Assign $v_i^\top$ as the $i^{th}$ row of $V$. 
      Assign $V_{-i} = \begin{bmatrix} 
      v_1 & \cdots & v_{i-1} & v_{i+1} & \cdots & v_n \end{bmatrix}^\top$.\;
      Solve the LASSO problem 
      $c_i = \arg\min_{\beta} 
      \frac{1}{2} ||v_i - V_{-i} \beta||_2^2 + \lambda ||\beta||_1$.\;
      Assign $\tilde{c}_i = \begin{bmatrix} 
      c_i^{(1)} & \cdots & c_i^{(i-1)} & 0 & c_i^{(i)} & \cdots & c_i^{(n-1)}
      \end{bmatrix}^\top$ such that the superscript is the index of 
      $\tilde{c}_i$.\;
    }
    Assign 
    $C = \begin{bmatrix} \tilde{c}_1 & \cdots & \tilde{c}_n \end{bmatrix}$.\;
    Compute the affinity matrix $B = |C| + |C^\top|$.\;
    Construct graph $G$ using $B$ as its similarity matrix.\;
    Partition $G$ into $K$ disconnected subgraphs (e.g., using edge 
    thresholding or spectral clustering).\;
    Map each partition to the community labels $1, ..., K$. 
\end{algorithm}

\begin{theorem}
\label{theorem5}
Let $P_n$ describe the edge probability matrix of the PABM with 
$n$ vertices, and let $A_n \sim \Bernoulli(P_n)$.  Let $\hat{V}_n$ be the 
matrix of eigenvectors of $A_n$ corresponding to the $K (K + 1) / 2$ most
positive and $K (K - 1) / 2$ most negative eigenvalues. Then 
$\exists \lambda > 0$ such that $\hat{B}_n$, the affinity matrix
outputted by SSC using $\hat{V}_n$ and $\lambda$, has $ij^{th}$ entry
$\hat{B}_n^{(ij)} \stackrel{a.s.}{\to} 0$ $\forall i, j$ in different 
communities. 
\end{theorem}

@@

For any edge probability matrix $P$ for the PABM such that the rows and 
columns are organized by community, the $kl$^th^ block is an outer product 
of two vectors, i.e., $P^{(kl)} = \lambda^{(kl)} (\lambda^{(lk)})^\top$. 
Therefore, given $P^{(kl)}$, $\lambda^{(kl)}$ and $\lambda^{(lk)}$ are 
solvable exactly (up to multiplication by $-1$) using singular value 
decomposition. More specifically, let $P^{(kl)} = \sigma^2 u v^\top$ be the 
singular value decomposition of $P^{(kl)}$. $u \in \mathbb{R}^{n_k}$ and 
$v \in \mathbb{R}^{n_l}$ are vectors and 
$\sigma^2 > 0$ is a scalar. Then $\lambda^{(kl)} = \pm \sigma u$ and 
$\lambda^{(lk)} = \pm \sigma v$. 
Given the adjacency matrix $A$ instead of edge probability matrix $P$, we can 
simply use plug-in estimators (algorithm 3), which converge to the true 
parameters.

@@

\begin{theorem}
\label{theorem6}
Under regularity and sparsity assumptions, and under the further 
assumption that $K$ is fixed and community labels are known,

\begin{equation} \label{eq:thm6}
\max_{k, l \in \{1, ..., K\}} 
||\hat{\lambda}^{(kl)} - \lambda^{(kl)}|| = 
O_P \bigg(\frac{(\log n_k)^c}{\sqrt{n_k}} \bigg)
\end{equation}
\end{theorem}

@@

For each simulation, community labels are drawn from a multinomial 
distribution, the popularity vectors $\{\lambda^{(kl)}\}_K$ are drawn 
from two types of joint distributions depending on whether $k = l$, the edge
probability matrix $P$ is constructed using 
the popularity vectors, and finally an unweighted and undirected adjacency 
matrix $A$ is drawn from $P$. OSC is then used for community detection, 
and this method is compared against SSC \cite{noroozi2019estimation}
\cite{soltanolkotabi2014} and MM \cite{igraph}
\cite{307cbeb9b1be48299388437423d94bf1}. True 
community labels are used with Algorithm 3 to estimate the popularity 
vectors $\{\lambda^{(kl)}\}_K$, and this method is then compared against an
MLE-based estimator described in \citeauthor{noroozi2019estimation} and
\citeauthor{307cbeb9b1be48299388437423d94bf1}.

@@

Theorem \ref{theorem4} implies that OSC will result in not just in the error
rate converging to 0 but the error *count* as well.
We explore this in Fig. \ref{fig:clust_err_ct_sim}.
